Supporting Information: Thermal sensitivity across forest vertical profiles: patterns, mechanisms, and ecological implications

### 2022-02-17 vertical-thermal-review.Rmd
updated piponiot citation

## 3. Leaf traits 
<!--Lawren's paragraph for R2 comments on having older citations and connecting to newer work-->
Leaf traits vary vertically across forest strata at anatomical, structural and biochemical levels [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping $T_{leaf}$, leaf metabolism, and leaf thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a;  @zwienieckiHydraulicLimitationsImposed2004]. 
This vertical variation in leaf structure and composition has been long recognized, with sun and shade leaf traits distinguished for more than 100 years [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].
Early work attributed differences among leaves to sun versus shade [@wyliePrinciplesFoliarOrganization1951; @vogelSunLeavesShade1968], and this tendency grew given the parallel differences observed for plants grown in sun versus shade conditions [@boardmanComparativePhotosynthesisSun1977; @abramsLeafStructuralCharacteristics1990].
More recent work has included a focus not only on light, but also on leaf height in the canopy, as important in determining differences in mature leaf structure and function, as leaf traits may partially mitigate the effects of gravity and hydraulic pathlength that would affect leaves higher up in the canopy [@sackHowStrongIntracanopy2006; @kochLimitsTreeHeight2004; @burgessRegressionsLeafTraits2006].
For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
Thus, leaves develop differently throughout the canopy according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of the ontogenetic stage of the branches [Table `r table_leaf_traits`; @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
Plasticity throughout the canopy may emerge during new leaf development with leaves being influenced by information from adjacent mature leaves or apical meristem, including  (e.g., for deciduous species expanding leaves) a “memory” of previous conditions in that location, and may also acclimate during and after expansion to to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
Overall, leaves may optimize their biochemical, anatomical and structural traits according to their light environment and height (@buckleyOptimalCarbonPartitioning2021; @hikosakaOptimalNitrogenDistribution2014; @kitaoCanopyNitrogenDistribution2018; @niinemetsPhotosynthesisResourceDistribution2007). 
Yet, while these studies have clarified many details of leaf differences throughout the canopy, no papers have integrated the many differences in leaf traits throughout the canopy in a cohesive framework.
Rather, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species using sun leaves [@wrightWorldwideLeafEconomics2004, @keenanGlobalLeafTrait2016; @chenLeafEconomicsSpectrum2020]. 
As canopy shade is known to alter these trait relationships (@osnasDivergentDriversLeaf2018), further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.

Across the vertical gradient, traits vary (1) within individuals, across height and light gradients [e.g., @example_within-individual_references], (2) across individuals of the same species, often representing an ontogenetic trajectory from the understory to the canopy [e.g., @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a], and (3) across understory and canopy species [e.g., @example_interspecific_references]. 
The majority of studies characterizing variation in leaf traits or metabolism examine intraspecific patterns (categories 1 and 2), which are therefore the main focus of this review (Table `r table_leaf_traits`).
Generally, the pattern of fundamental trait adaptations to light gradients across forest species is similar to within-species trends in traits and sometimes metabolism [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @binLeafTraitExpression2022; @cavaleri_height_2010; @lloydOptimisationPhotosyntheticCarbon2010]; however, vertical trends vary across trees [@lloydOptimisationPhotosyntheticCarbon2010], and counter-gradient variation can exist [e.g., @luskWhyAreEvergreen2008].
Overall, when it comes to understanding and modeling forest ecosystem function, many leaf physiological parameters can be estimated based on foliage height profiles [e.g., @cavaleri_height_2010; @lloydOptimisationPhotosyntheticCarbon2010].

@@

Canopy trees account for the majority of forest ecosystem water and carbon cycling.
While studies partitioning transpiration across forest vertical gradients are rare, both evapo-transpiration (ET) and transpiration have been shown to increase with height in a *Picea abies* forest in Germany, such that the upper half of the canopy contributed an estimated 80% of daytime ET [@staudtVerticalStructureEvapotranspiration2011]. 
Similarly, in a tropical forest in the Brazilian Amazon, canopy and subcanopy trees jointly contributed approximately 93% of ET, or 88% of transpiration [Fig `r fig_ecosystem`a, @kunertRevisedHydrologicalModel2017], and trees >33 cm DBH contributed >80% of transpiration in a forest in Costa Rica [@mooreUpscalingTranspirationDiverse2018]. 
In terms of C cycling, it has been estimated that the canopy strata contributes $\ge$ 64% of net daytime CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil, @missonPartitioningForestCarbon2007]. 
Large trees also dominate in terms of woody above ground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$ 10cm DBH usually contributing > 85% of $ANPP_{woody}$ and $M_{woody}$ across 25 globally distributed forests [Fig. `r fig_ecosystem`b, @piponiotSizerelatedTreeBiomass2022].

![**Figure `r fig_ecosystem`. Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests.** Panel (a) is from @kunertRevisedHydrologicalModel2017; panel (b) presents averages for 14 tropical forests from @piponiotSizerelatedTreeBiomass2022.](schematics/ecosystem vertical partitioning/ecosystem partitioning.png)

>>>>

## 3. Leaf traits 
<!--Lawren's paragraph for R2 comments on having older citations and connecting to newer work-->
Leaf traits vary vertically across forest strata at anatomical, structural and biochemical levels [Table `r table_leaf_traits`, @sackHowStrongIntracanopy2006; @niinemetsWorldwideAnalysisWithincanopy2015; @chenLeafEconomicsSpectrum2020], shaping $T_{leaf}$, leaf metabolism, and leaf thermal sensitivity across forest vertical strata [@michaletzPlantThermoregulationEnergetics2015a;@michaletzEnergeticCarbonEconomic2016a; @zwienieckiHydraulicLimitationsImposed2004]. 
This vertical variation in leaf structure and composition has been long recognized, with sun and shade leaf traits distinguished for more than 100 years [@haberlandtPhysiologicalPlantAnatomy1914; @salisburyCausesEcologicalSignificance1928].
Early work attributed differences among leaves to sun versus shade [@wyliePrinciplesFoliarOrganization1951; @vogelSunLeavesShade1968], and this tendency grew given the parallel differences observed for plants grown in sun versus shade conditions [@boardmanComparativePhotosynthesisSun1977; @abramsLeafStructuralCharacteristics1990].
More recent work has included a focus not only on light, but also on leaf height in the canopy, as important in determining differences in mature leaf structure and function, as leaf traits may partially mitigate the effects of gravity and hydraulic pathlength that would affect leaves higher up in the canopy [@sackHowStrongIntracanopy2006; @kochLimitsTreeHeight2004; @burgessRegressionsLeafTraits2006].
For example, leaf mass per area (LMA), photosynthetic capacity ($A_{max}$ per area) and $g_s$ have been observed to increase with height, independently of light [@thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @cavender-baresChangesDroughtResponse2000; @binLeafTraitExpression2022].
Thus, leaves develop differently throughout the canopy according to the irradiance and hydraulic stress associated with their canopy location, height and evaporative load, with additional potential influences of the ontogenetic stage of the branches [Table `r table_leaf_traits`; @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a; @casasSunShadeLeaves2011;  @keenanGlobalLeafTrait2016; @binLeafTraitExpression2022; @sackHowStrongIntracanopy2006; @chenLeafEconomicsSpectrum2020; @niinemetsVariabilityLeafMorphology2015].
Plasticity throughout the canopy may emerge during new leaf development with leaves being influenced by information from adjacent mature leaves or apical meristem, including  (e.g., for deciduous species expanding leaves) a “memory” of previous conditions in that location, and may also acclimate during and after expansion to to the current microclimate [@zwienieckiHydraulicLimitationsImposed2004].
Overall, leaves may optimize their biochemical, anatomical and structural traits according to their light environment and height (@lloydOptimisationPhotosyntheticCarbon2010; @buckleyOptimalCarbonPartitioning2021; @hikosakaOptimalNitrogenDistribution2014; @kitaoCanopyNitrogenDistribution2018; @niinemetsPhotosynthesisResourceDistribution2007). 
Yet, while these studies have clarified many details of leaf differences throughout the canopy, no papers have integrated the many differences in leaf traits throughout the canopy in a cohesive framework.
Rather, much of our understanding of trait coordination is based on the leaf economics spectrum concept developed across diverse species using sun leaves [@wrightWorldwideLeafEconomics2004, @keenanGlobalLeafTrait2016; @chenLeafEconomicsSpectrum2020]. 
As canopy shade is known to alter these trait relationships (@osnasDivergentDriversLeaf2018), further research is needed to characterize trait relationships and responses vertically through the full range of canopy microenvironments.

Across the vertical gradient, traits vary (1) within individuals, across height and light gradients [e.g., @zwienieckiHydraulicLimitationsImposed2004; @sackHowStrongIntracanopy2006; @binLeafTraitExpression2022], (2) across individuals of the same species, often representing an ontogenetic trajectory from the understory to the canopy [e.g., @carterExperimentalWarmingTropical2021; @niinemetsResponsesForestTrees2010a], and (3) across understory and canopy species [e.g., @lloydOptimisationPhotosyntheticCarbon2010]. 
The majority of studies characterizing variation in leaf traits or metabolism examine intraspecific patterns (categories 1 and 2), which are therefore the main focus of this review (Table `r table_leaf_traits`).
Generally, the pattern of fundamental trait adaptations to light gradients across forest species is similar to within-species trends in traits and sometimes metabolism [@rozendaalPlasticityLeafTraits2006; @thomasPhotosyntheticDifferencesSaplings2002; @houterOntogeneticChangesLeaf2012; @binLeafTraitExpression2022; @cavaleri_height_2010; @lloydOptimisationPhotosyntheticCarbon2010]; however, vertical trends vary across trees [@lloydOptimisationPhotosyntheticCarbon2010], and counter-gradient variation can exist [e.g., @luskWhyAreEvergreen2008].
Overall, when it comes to understanding and modeling forest ecosystem function, many leaf physiological parameters can be estimated based on foliage height profiles [e.g., @cavaleri_height_2010; @lloydOptimisationPhotosyntheticCarbon2010].

@@

Canopy trees account for the majority of forest ecosystem water and carbon cycling.
While studies partitioning transpiration across forest vertical gradients are rare, both evapo-transpiration (ET) and transpiration have been shown to increase with height in a *Picea abies* forest in Germany, such that the upper half of the canopy contributed an estimated 80% of daytime ET [@staudtVerticalStructureEvapotranspiration2011]. 
Similarly, in a tropical forest in the Brazilian Amazon, canopy and subcanopy trees jointly contributed approximately 93% of ET, or 88% of transpiration [Fig `r fig_ecosystem`a, @kunertRevisedHydrologicalModel2017], and trees >33 cm DBH contributed >80% of transpiration in a forest in Costa Rica [@mooreUpscalingTranspirationDiverse2018]. 
In terms of C cycling, it has been estimated that the canopy strata contributes $\ge$ 64% of net daytime CO~2~ uptake [i.e., GPP - ecosystem respiration, including from soil, @missonPartitioningForestCarbon2007]. 
Large trees also dominate in terms of woody above ground net primary productivity ($ANPP_{woody}$) and mortality ($M_{woody}$), with trees $\geq$ 10cm DBH usually contributing > 85% of $ANPP_{woody}$ and $M_{woody}$ across 25 globally distributed forests [Fig. `r fig_ecosystem`b, @piponiotDistributionBiomassDynamicsinpress].

![**Figure `r fig_ecosystem`. Vertical partitioning of (a) evapotranspiration and (b) C fluxes in tropical forests.** Panel (a) is from @kunertRevisedHydrologicalModel2017; panel (b) presents averages for 14 tropical forests from @piponiotDistributionBiomassDynamicsinpress.](schematics/ecosystem vertical partitioning/ecosystem partitioning.png)
